% To support generating a plain text version, this file can contain only plaintext,
% comments, and the following macros:
% - \emph
Dependently typed languages have proven useful for developing large-scale fully
verified software, but we do not have any guarantees after compiling that
verified software.
A verified program written in a dependently typed language, such as Coq, can be
type checked to ensure that the program meets its specification.
Similarly, type checking prevents us from importing a library and
violating the specification declared by its types.
Unfortunately, we cannot perform either of these checks after compiling a
dependently typed program, since all current implementations erase types before
compiling the program.
Instead, we must trust the compiler to not introduce errors into the
verified code, and, after compilation, trust the programmer to never introduce
errors by linking two incompatible program components.
As a result, the compiled and linked program is
\emph{not verified}---we have no guarantees about what it will do.

In this dissertation, I develop a theory for preserving dependent types through
compilation so that we can use type checking after compilation to check that no
errors are introduced by the compiler or by linking.
Type-preserving compilation is a well-known technique that has been used to
design compilers for non-dependently typed languages, such as ML, that
statically enforce safety and security guarantees in compiled code.
But there are many open challenges in scaling type preservation to dependent
types.
The key problems are adapting syntactic type systems to interpret low-level
representations of code, and breaking the complex mutually recursive structure
of dependent type systems to make proving type preservation and compiler
correctness feasible.
In this dissertation, I explain the concepts required to scale type preservation
to dependent types, present a proof architecture and language design that
support type preservation, and prove type preservation and compiler correctness
for four early-stage compiler translations of a realistic dependently typed
calculus.
These translations include an A-normal form (ANF), a continuation-passing style
(CPS), an abstract closure conversion, and a parametric closure conversion
translation.
