\pdfbookmark[1]{Acknowledgments}{Acknowledgments}

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgments}
I never thought I would get a Ph.D.
In fact, I thought I would never get a Ph.D.
As I completed my undergraduate degree, a Ph.D. seemed like an insurmountable
challenge: I would have to produce knowledge unknown to the world, write about,
talk about it, and teach it.
I didn't know how to do that.
I didn't want to do that.
I just wanted to write fun programs and learn interesting facts.

I was wrong.
The following people helped me realize that.
\begin{itemize}
\item Dan Friedman

  Dan is the reason I'm here today.

  Dan introduced me to the beauty of programming languages, causing me to shift
  gears from software engineering and security to spending my free time reading
  about language design.
  When I was interested in research, Dan recommended I talk to Amal.
  He explicitly recommended that I not work with him, claiming that he doesn't
  do important research.
  That might be, but he does do important work.
\item Amr Sabry

  Amr taught me how to write software, and gave me a few research problems to
  play with as a Master's student while I tried to figure out whether I wanted
  to do a Ph.D.
\item R. Kent Dybvig

  Kent taught me that most programs aren't written by humans, and how to think
  like a machine.
\item The Northeastern PRL, jointly and separately.

  Vincent St-Amour, Asumu Takikawa, Phillip Mates, Jamie Perconti, Paul
  Stansifer, Tony Garnock-Jones, Stephen Chang, Justin Slepak, Jonathan
  Schuster, Dionna Amalie Glaze, Andrew Cobb, Sam Caldwell, Ben Greenman, Leif
  Andersen, Daniel Patterson, Aaron Weiss, and Ben Lerner.

  Thank you all for hanging out at happy hours, criticizing my practice talks,
  and making my Ph.D. a great personal and professional experience.
\item Max S. New

  Max, the category practitioner.
  He understands abstraction in ways I doubt I ever will.
  Many times, he has helped me understand my own work better.
  Max is a brilliant friend and amazing colleague.
\item Cătălin Hriţcu

  Cătălin is a kind, generous, and skeptical person---a great scientist.
  He is one of the first people I met at a programming languages conference.
  Cătălin hosted me at Inria for a semester while I finished this work, giving
  me the chance to meet people who know way more about dependent types than me.
  This dissertation is much better because of that experience.
\item Annabel Satin

  Annabel is an awesome person, working tirelessly in the background for the
  good of programming languages.
  Without her, our work might well grind to a halt.
\item Youyou Cong

  Youyou visited us for a semester, and was critical to part of this
  dissertation.
  She has an encyclopedic knowledge of the literature surrounding CPS.
  I still sometimes bother her with my ideas.
\item Neel Krishnaswami

  Every time I talk to Neel, I learn something awesome about semantics.
  Our chats have shaped my work in subtle ways.
\item Ron Garcia

  Ron is the nicest person in the programming languages community.
  Talking with Ron, I feel like he is genuinely interested in what I have to
  say, even when all I have to say are half baked ideas.
  He can express his confusion, or clearly explain why I'm wrong, without
  making me feel stupid.
\item Éric Tanter

  Éric has always made me feel like a colleague, and not just a graduate student.
  He's one of the reasons I'm sticking around in academia.
\item Stephanie Weirich

  Stephanie has a unique perspective on dependent types.
  She has challenged my assumptions and design choices throughout my
  dissertation.
  One day, I'll try to put those thoughts in writing.
\item Greg Morrisett

  Greg has been a huge help during this work.
  Even after starting as dean, Greg found time to chat about research.
  When I sent him a draft, he'd read it in mere hours and come back with cogent
  feedback, questions, and criticisms.
\item Mitch Wand

  Mitch helped me learn more about writing and English during this dissertation
  than I learned in 22 years of school before I started.
\item Matthias Felleisen

  Matthias has constantly pushed me to reach a higher standard for scientific
  rigor, precision, and clarity---more so than anyone else.
  He has never hesitated to (constructively) criticize my work.
  I've learned a ton from him, and if I'm lucky I'll learn some more.
\item Amal Ahmed

  Amal has been a great advisor.
  She has absolute and infectious enthusiasm for what she does.
  She helped me learn about proofs and types, and about communicating complex
  mathematical work.
  She chewed me out when I ignored important things, and gave me latitude
  when I needed it.

  But mostly, Amal taught me to not fear hard problems.
  In 2010, as an undergrad, I asked Amal if she had any research problems.
  She gave me a book (Types and Programming Languages), and a secure compilation
  problem, and said it would take ``6 months, max''.
  In 2015, we finished that paper.
  In 2016, I had a choice: continue working on secure compilation, or switch to
  working on dependent types.
  I was more interested in dependent types, but switching topics 4 years in to a
  Ph.D. program seemed like a big risk, so I chose secure compilation.
  A few days later, I walked into her office and said ``I changed my mind. It's
  going to be hard, but it'll be way more interesting.''.
  I don't know if I could have done that without Amal.
\end{itemize}

\bigskip

\noindent Thank you all. You have all shaped this work, either directly or indirectly through emotional and psychological support, and are part of the reason I finished this dissertation. And thank you to many more people who I've not put in this list; there are many of you who have helped me in ways I haven't been able to put into words.

\hfill --- William J. Bowman
\endgroup
